Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-13 csf-2017/paper.tex
Merge a discussion by Lars into a related work discussion that I accidentally rewrote without checking what was already there...

\subsection{Discussion}
%/comparison with related Dreyer-Neis-Birkedal and with Hur-Dreyer}
%\todo[inline]{Add discussion/comparison to other LRs}
% Comparisson of LR to other work:
% - Single othorgonal closure
%   o all capabilities are possible continuations, not just the one we
%   intend to be it.
% - No termination guarantees in observation relation
% - Private/public transitions used like in Dreyer-Neis-Birkedal, but
%   also to  handle the difference in what local and global
%   capabilities can handle.

\todo[inline]{Maybe move this subsection to related work}

As already mentioned, our the state transition systems used in our
logical relation are inspired by earlier work on logical relations for
high-level
languages~\cite{Ahmed:popl09, Dreyer:jfp12,Devriese:2016ObjCap}.
The idea of public and private transitions comes from 
Dreyer et. al.~\cite{Dreyer:jfp12}, who gave a logical relations model
of a high-level lambda calculus. In \emph{loc. cit.}
public transitions were used to 
capture the \emph{extensional} behavior of a program expression
(and a context), whereas private transitions were used to capture
the internal behaviour of a computation using local state.
Technically, the logical relation of Dreyer et. al.\ was
formulated using biorthogonal closure, and the future world
relation for continuations was required to be a public future world
relation. In other words, the choice between private or public future
rested on the ``structure'' of the programming language. 
In contrast, 
for our low-level, unstructured, assembly language, we
we instead decide on whether to use a private or public future 
world depending on whether the capability is global or local. 
Another interesting technical difference is that in our low-level 
language we do not have a fixed calling convention and any register
can serve as a continuation, and therefore we do not use biorthogonal
closure but just model return pointers as continuations using a single
orthogonal closure.

\lars{don't know yet whether we need something like the following paragraph:}
Even though 
Hur and Dreyer~\cite{Hur:2011:KLR:1926385.1926402} also
considered a logical
relation involving an assembly language, they still used
biorthogonal closure but that was because their logical relation
was a relation between a high-level language and an assembly
language,  and thus the assembly language in the relation essentially
inherited structure from the high-level language.

@@

\item Some compartmentalisation result?
\end{itemize}
\end{comment}
\section{Discussion, future work}
\label{sec:discussion}
\begin{itemize}
\item A general well-bracketed control flow result?
\begin{itemize}
\item what would that result say?
\item possible idea: fully abstract compilation from an assembly language with
a trusted stack to one without
\item the LR and some of the lemmas already imply well-bracketed control flow, as seen in examples
\end{itemize}
\item Relation to local parameters in Scala, Algol/Pascal second-class function parameters?
\item Stack clearing realistic?
\item Non-modularity of heap allocation requirement for adversary callbacks
\end{itemize}

@@

Our logical relation is a unary, step-indexed Kripke logical relation (SKLR)
with recursive
worlds~\cite{Appel:2001:IMR:504709.504712,Ahmed2004semantics,Birkedal:2011:SKM:1926385.1926401},
closely related to the one used by \citet{Devriese:2016ObjCap} to formulate
capability-safety in a high-level JavaScript-like lambda calculus. Our
fundamental theorem is similar to theirs and can be seen as a formulation of
capability-safety of the capability-machine. Because we are not interested in an
externally observable notion of side-effects (like console output or memory
access traces), we do not require their notion of effect parametricity.

The idea of applying SKLRs to an assembly language was applied before by
\citet{Hur:2011:KLR:1926385.1926402}, who constructed a binary SKLR for
formulating and proving correctness of an ML-to-assembly compiler. Their
condition on the assembly code are quite loose (for example, they allow
hand-written and self-modifying assmebly code), but they do assume code that
is non-adversarial and respects the ABI of the high-level language (for example,
the code does not store return pointers and invoke them after their intended
lifetime). As a result, they can treat return pointers as evaluation contexts in
a lambda calculus and use biorthogonality~\cite{pitts_operational_1998}. This
means that they define the logical relation to accept evaluation contexts if
they produce related observations for related values and to accept terms if they
produce related observations when plugged into related evaluation contexts. In
our setting, there is no guarantee that adversarial code will treat return
pointers as intended, so, as described in Section~\ref{sec:logical-relation}, we
define our logical relation in a (arguably natural but still) novel way, using a single
orthogonal closure, i.e. accept terms if they produce related observations when
given related arguments, where those arguments include the return pointer.

Another novel aspect of our logical relation is how we model the temporary
nature of local capabilities. The main insight is that the temporary, revokable
nature of these capabilities can be seen as a generalisation of the special
status of evaluation contexts in lambda calculi. Closely related to this is a
study by \citet{Dreyer:jfp12} (DNB) of the impact of higher-order state and
control effects on local relational reasoning. DNB formally capture this special
status of evaluation contexts using Kripke worlds with public and private future
world relations. Our public and private worlds are inspired by theirs and they
are at times remarkably similar. For example, the state-transitition system in
the proof of our assembly version of the awkward example, is exactly the same as
the one used by DNB for the original awkward example. Nevertheless, there are
also remarkable differences: in DNB's logical relation, public future worlds
only turn up in the continuation relation (which we do not have, see above).
This is different from our value relation which uses public and private future
worlds depending on the locality of the capability at hand. Additionally, our
worlds are a bit more complex because, to allow store-local capabilities, they
can have temporary regions that are only monotonous w.r.t. public future worlds,
while DNB's worlds are entirely permanent.

As a side note: our model of local/global capabilities may also be applicable
for reasoning about related ideas in higher-level languages. Specifically,
\citet{osvald_gentrification_2016}'s second-class or local values in Scala are
remarkably similar to the local capabilities we study and we expect they will
support similar reasoning patterns.

>>>>

@@

\item Some compartmentalisation result?
\end{itemize}
\end{comment}

\section{Discussion, future work}
\label{sec:discussion}

\begin{itemize}
\item A general well-bracketed control flow result?
  \begin{itemize}
  \item what would that result say?
  \item possible idea: fully abstract compilation from an assembly language with
  a trusted stack to one without
  \item the LR and some of the lemmas already imply well-bracketed control flow, as seen in examples
  \end{itemize}
\item Relation to local parameters in Scala, Algol/Pascal second-class function
  parameters? 
\item Stack clearing realistic?
\item Non-modularity of heap allocation requirement for adversary callbacks
\end{itemize}

It is worth mentioning that local capabilities are quite similar to features
that have been proposed for high-level languages. Specifically,
\citet{osvald_gentrification_2016}'s second-class or local values in Scala are
remarkably similar and we expect they will support similar reasoning patterns.

@@

Our logical relation is a unary, step-indexed Kripke logical relation (SKLR)
with recursive
worlds~\cite{Appel:2001:IMR:504709.504712,Ahmed2004semantics,Birkedal:2011:SKM:1926385.1926401},
closely related to the one used by \citet{Devriese:2016ObjCap} to formulate
capability-safety in a high-level JavaScript-like lambda calculus. Our
fundamental theorem is similar to theirs and can be seen as a formulation of
capability-safety of the capability-machine. Because we are not interested in
externally observable side-effects (like console output or memory access
traces), we do not require their notion of effect parametricity.

The idea of applying SKLRs to an assembly language was applied before by
\citet{Hur:2011:KLR:1926385.1926402}, who constructed a binary SKLR for
formulating and proving correctness of an ML-to-assembly compiler. Their
condition on the assembly code are quite loose (for example, they allow
hand-written and self-modifying assembly code), but they do assume code that is
non-adversarial and respects the calling convention of the high-level language
(for example, the code does not store return pointers and invoke them after
their intended lifetime). As a result, they can treat return pointers as
evaluation contexts (continuations) in a lambda calculus and use biorthogonality
to define the LR~\cite{pitts_operational_1998}. In our setting, there is no
guarantee that potentially adversarial code will respect the calling convention,
so, as described in Section~\ref{sec:logical-relation}, we do not use
biorthogonal closure but just model return pointers as continuations using a
single orthogonal closure.

Another novel aspect of our logical relation is how we model the temporary
nature of local capabilities. The main insight is that the temporary, revokable
nature of these capabilities can be seen as a generalisation of the
syntactically special, unstoreable status of evaluation contexts in lambda
calculi (without control effects). Closely related to this is a study by
\citet{Dreyer:jfp12} (DNB) of the impact of higher-order state and control
effects on local relational reasoning. DNB formally capture the special status
of evaluation contexts using Kripke worlds with public and private future world
relations. Our future world relations are inspired by theirs and they are at
times remarkably similar. For example, the state-transitition system in the
proof of our assembly version of the awkward example, is exactly the same as the
one used by DNB for the original awkward example. However, in our logical
relation, public and private future worlds turn up in an entirely different
place in the logical relation: rather than using public future worlds for the
special syntactic category of evaluation contexts, they are used in the value
relation depending on the depending on the locality of the capability at hand.
Additionally, our worlds are a bit more complex because, to allow store-local
capabilities, they can have temporary regions that are only monotonous w.r.t.
public future worlds, while DNB's worlds are entirely permanent.

%/comparison with related Dreyer-Neis-Birkedal and with Hur-Dreyer}
%\todo[inline]{Add discussion/comparison to other LRs}
% Comparisson of LR to other work:
% - Single othorgonal closure
%   o all capabilities are possible continuations, not just the one we
%   intend to be it.
% - No termination guarantees in observation relation
% - Private/public transitions used like in Dreyer-Neis-Birkedal, but
%   also to  handle the difference in what local and global
%   capabilities can handle.

% As already mentioned, our the state transition systems used in our
% logical relation are inspired by earlier work on logical relations for
% high-level
% languages~\cite{Ahmed:popl09, Dreyer:jfp12,Devriese:2016ObjCap}.
% The idea of public and private transitions comes from 
% Dreyer et. al.~\cite{Dreyer:jfp12}, who gave a logical relations model
% of a high-level lambda calculus. In \emph{loc. cit.}
% public transitions were used to 
% capture the \emph{extensional} behavior of a program expression
% (and a context), whereas private transitions were used to capture
% the internal behaviour of a computation using local state.
% Technically, the logical relation of Dreyer et. al.\ was
% formulated using biorthogonal closure, and the future world
% relation for continuations was required to be a public future world
% relation. In other words, the choice between private or public future
% rested on the ``structure'' of the programming language. 
% In contrast, 
% for our low-level, unstructured, assembly language, we
% we instead decide on whether to use a private or public future 
% world depending on whether the capability is global or local. 
% Another interesting technical difference is that in our low-level 
% language we do not have a fixed calling convention and any register
% can serve as a continuation, and therefore we do not use biorthogonal
% closure but just model return pointers as continuations using a single
% orthogonal closure.

% \lars{don't know yet whether we need something like the following paragraph:}
% Even though 
% Hur and Dreyer~\cite{Hur:2011:KLR:1926385.1926402} also
% considered a logical
% relation involving an assembly language, they still used
% biorthogonal closure but that was because their logical relation
% was a relation between a high-level language and an assembly
% language,  and thus the assembly language in the relation essentially
% inherited structure from the high-level language.
